---
title: "Untitled"
output: html_document
date: "2024-11-19"
---

```{r}
setwd("/Users/jacobrichards/Desktop/DS_DA_Projects/Project_3")
data <- read.csv(file="lease_renewal.csv",header=TRUE)
data
```

Important Note: Please do not build any kind of regression or classification model to address the questions below

What are the key drivers that explain whether a resident will renew the lease or not? In other words, what are the renewal promoters and detractors?

First lets divide the data set into tenants that renewed and those that did not.

```{r}
did_renew <- data[data[,18] == 1, ]
did_not_renew <- data[data[,18] == 0, ]

#whats the overall % of non-renewal

(nrow(did_not_renew)/nrow(data))

#late payments equals one means that the tenant has never been late on making a payment.


```

lets check the effect of rent change

no_rent_change: 0 if rent change was less than 10% COL 2

rent_change_10: 0 if was 10% COL 3

rent_change_20: 0 if was 20% COL 4

of those who's rent changed less than 10% but still changed what ratio of renewed

```{r}
less_10_rent_changed <- data[data[,2] == 0,]


less_10_rent_changed_renewed <- less_10_rent_changed[less_10_rent_changed[,18] == 1, ]
less_10_rent_changed_did_not_renewed <- less_10_rent_changed[less_10_rent_changed[,18] == 0, ]

cat("Percentage that did renew:", nrow(less_10_rent_changed_renewed) / nrow(less_10_rent_changed), "\n")
cat("Percentage that did not renew:", nrow(less_10_rent_changed_did_not_renewed) / nrow(less_10_rent_changed), "\n")



x <- numeric(2)
x[1]<- nrow(less_10_rent_changed_did_not_renewed)
x[2]<- nrow(did_not_renew)

n <- numeric(2)
n[1] <- nrow(less_10_rent_changed)
n[2] <- nrow(data)

p <- numeric(2)
p[1] <- nrow(less_10_rent_changed_did_not_renewed)/nrow(less_10_rent_changed)
p[2] <- nrow(did_not_renew)/nrow(data)

library(epitools)



```

```         
of those that did have a rent increase less than 10%, 0.1668356 renewed
and 0.8331644 did not renew. 
```

lets check the effect of lease length

```{r}

```

lets check the effect of age

```{r}

```
