---
title: "Module 12 Homework"
output: html_document
date: "2024-11-08"
---
1.) (Use R) Use the mpg data from ggplot2 package and create R codes necessary to
generate the following graphs.
```{r}
library(ggplot2)
data(mpg)
```

a.)
```{r}
ggplot(data=mpg, aes(x=displ,y=hwy)) + geom_point() + geom_smooth() + xlim(1.75,7) + ylim(10,45)
```

b.)
```{r}
ggplot(data=mpg, aes(x=displ,y=hwy, color=drv)) + geom_point() + geom_smooth(se=FALSE) + xlim(1.75,7) + ylim(10,45)
```

c.)
```{r}
ggplot(data=mpg, aes(x=displ, y=hwy)) + geom_point(aes(color=drv)) + geom_smooth(se = FALSE) + xlim(1.75, 7) + ylim(10, 45)
```

d.)
```{r}
ggplot(data=mpg, aes(x=displ, y=hwy, color=drv, linetype=drv)) + geom_point() + 
  geom_smooth(data = subset(mpg, drv == "4"), se = FALSE, color = "blue") + 
  geom_smooth(data = subset(mpg, drv == "f"), se = FALSE, color = "blue") + 
  geom_smooth(data = subset(mpg, drv == "r"), se = FALSE, color = "blue") + xlim(1.75, 7) + ylim(10, 45) + 
  scale_linetype_manual(values = c("4" = "solid", "f" = "dotted", "r" = "dotted")) 
```

2.) (Use R) The iris is a data frame with 150 cases (rows) and 5 variables (columns)
named Sepal.Length, Sepal.Width, Petal.Length, Petal.Width, and Species. Use the iris
data and create R codes necessary to generate the following graphs.
a.)
```{r}
data(iris)
ggplot(data = iris, aes(x = Petal.Length, y = Petal.Width, color = Species, shape = Species)) + 
  geom_point() + scale_shape_manual(values = c("setosa" = 16, "versicolor" = 17, "virginica" = 15)) + 
  labs(x = "Petal Length", y = "Petal Width", title = "Scatter plot")
```

b.)
```{r}
ggplot(data = iris, aes(x = Petal.Length, y = Petal.Width, color = Species, shape = Species)) + 
  geom_point() + labs(x = "Petal Length", y = "Petal Width", title = "Scatter plot with linear Model") +
  geom_smooth(method="lm")
```

c.)
```{r}
ggplot(data = iris, aes(x = Petal.Length, y = Petal.Width, color = Species, shape = Species)) + 
  geom_point() + labs(x = "Petal Length", y = "Petal Width", title = "Scatter plot with local polynomial regression") +
  geom_smooth(method = "loess")
```


```{r}
ggplot(data = iris, aes(x = Petal.Length, y = Petal.Width, color = Species)) + 
  geom_line(aes(group = Species)) +  
  labs(x = "Petal Length", y = "Petal Width", title = "Scatter plot with local polynomial regression") 
```

e.) 
```{r}
ggplot(data = iris, aes(x = Petal.Length, y = Petal.Width, color = Species, shape = Species)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  labs(x = "Petal Length", y = "Petal Width", title = "Faceting") +
  facet_wrap(vars(Species), scales = "free") + ylim(0,2.6)
```

f.)
```{r}
ggplot(data = iris, aes(x = Petal.Length, y = Petal.Width, color = Species, shape = Species)) + 
  geom_point() + labs(x = "Petal Length", y = "Petal Width", title = "Faceting") +
  geom_smooth(method = "loess") + facet_wrap(vars(Species), scales = "free")
```

g.)
```{r}
ggplot(data = iris, aes(x = Species, y = Petal.Length, fill = Species)) + 
  geom_boxplot(position = "dodge") + 
  labs(y = "Petal Length", title = "Box Plot") 
```

h.) 
```{r}
ggplot(data = iris, aes(x = Species, y = Petal.Length, fill = Species)) + 
  geom_violin(position = "dodge") + 
  labs(y = "Petal Length", title = "Violin Plot")
```

i.)
```{r}
ggplot(data = iris, aes(x = Species, y = Petal.Length, fill = Species)) + 
  geom_violin(trim = FALSE, position = "dodge") + 
  geom_boxplot(width = 0.1, position = position_dodge(width = 0.9), fill = "orange") +
  labs(y = "Petal Length", title = "Violin/Box") 
```

j.)
```{r}
library(ggridges)
ggplot(data = iris, aes(x = Petal.Length, y = Species, fill = Species)) + 
  geom_hline(yintercept = c(1, 2, 3), color = "black", size = .6) +  # Adds thick horizontal lines for each Species level
  geom_density_ridges(scale = 1.75, rel_min_height = 0.01) + 
  labs(x = "Petal Length",y="frequency", title = "Ridgeline Plot of Petal Length by Species") + 
  theme_bw() + 
  theme(
    plot.title = element_text(face = "bold", size = 12), 
    legend.background = element_rect(fill = "white", linewidth = 2, colour = "white"),
    panel.grid.minor = element_blank()
  ) +   
  expand_limits(y = c(0.5, 3.75))  # Adds a bit of space around the plot

```

k.)
```{r}
library(ggplot2)
species_counts <- as.data.frame(table(iris$Species))
colnames(species_counts) <- c("Species", "Count")
ggplot(species_counts, aes(x = "", y = Count, fill = Species)) + 
  geom_bar(stat = "identity", width = 1, color = "black") +  
  coord_polar("y") + 
  theme_void()
```

l.)
```{r}
library(ggplot2)
ggplot(iris, aes(Petal.Length, fill = Species)) + 
geom_bar(position = "stack", color = "black", size = 0.25) + 
labs(y = "Frequency", title = "Histogram")


```

m.) 
```{r}
library(ggplot2)
scaling_factor <- nrow(iris) * diff(range(iris$Petal.Length)) / 55  
ggplot(iris, aes(x = Petal.Length)) + 
  geom_bar(fill = "red", color = "black", binwidth = .10) +  
  geom_density(aes(y = ..density.. * scaling_factor), size = 1, color = "blue",fill="blue",alpha=0.15) + 
  labs(y = "Frequency", title = "Histogram of Petal Length with Density Curve") +
  theme_minimal()
```

n.) 
```{r}
ggplot(iris, aes(x = Petal.Length, fill = Species, color = Species)) + 
  geom_density(alpha = 0.4) +  # Add semi-transparent density curves for each species
  labs(x = "Petal Length", y = "Density", title = "Kernel Density Plot of Petal Length by Species") +
  theme_minimal()
```

3.) (Use R) Consider the ‘College Rankings 2014’ dataset. Use any graphing or plotting
techniques to glean some insight from the data. At the very least include an interesting plot
of each of the following:
a. Two categorical variables
b. Two quantitative variables
c. A categorical and quantitative variable
d. Any single variables of interest.
e. Any other interesting plots you may find.


```{r}
setwd("/Users/jacobrichards/Desktop/Stat_405_R/Module 12")
college_rankings <- read.csv(file="College_Rankings_2014.csv",header=TRUE)
```

a. Two categorical variables
```{r}
counts <- table(college_rankings$Type)
totall <- sum(counts)
Type_percentage <- as.data.frame(counts / totall)
Type_percentage$Group <- "Type"  

state_counts <- table(college_rankings$State)
totall_totall <- sum(state_counts)
state_percentages <- as.data.frame(state_counts / totall_totall)
state_percentages$Group <- "State"  

concontenate <- rbind(Type_percentage, state_percentages)

concontenate$Facet_Group <- ifelse(concontenate$Group == "Type", "Type (Private/Public)", "State Distribution")

categorical <- ggplot(concontenate, aes(x = Var1, y = Freq, fill = Freq)) + 
  geom_bar(stat = "identity", color = "black") +
  labs(title = "Percentage of Private vs Public Colleges and State Distribution", y = "Percentage", x = "Category") + 
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 3, angle = 45, hjust = 1)  # Sets x-axis text to a small size
  ) +
  facet_wrap(~ Facet_Group, scales = "free") 
```

b. Two quantitative variables
```{r}
library(ggplot2)
library(tidyr)
conc <- cbind(college_rankings$Acceptance.rate,college_rankings$X6yr.Grad.rate)
quant <- as.data.frame(conc)

quant_long <- quant %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value")

ggplot(quant_long, aes(x = Value, fill = Variable, color = Variable)) + 
  geom_density(alpha = 0.4) + 
  labs(title = "Distribution of Each Variable in Quant", x = "Value", y = "Density of rate") +
  theme_minimal() +
  xlim(-15, 130) +
  scale_fill_manual(values = c("V1" = "blue", "V2" = "red"),
                    labels = c("V1" = "Acceptance Rate", "V2" = "6-Year Graduation Rate")) +
  scale_color_manual(values = c("V1" = "blue", "V2" = "red"),
                     labels = c("V1" = "Acceptance Rate", "V2" = "6-Year Graduation Rate"))

```

c. A categorical and quantitative variable
```{r}
Public_enrollment <- college_rankings[college_rankings$Type == "Public", "Enrollment"]
Private_enrollment <- college_rankings[college_rankings$Type == "Private", "Enrollment"]
enrollment_data <- data.frame(Enrollment = c(Public_enrollment, Private_enrollment),
Type = c(rep("Public", length(Public_enrollment)), rep("Private", length(Private_enrollment))))
ggplot(enrollment_data, aes(x = Enrollment, fill = Type, color = Type)) + 
  geom_density(alpha = 0.4) + labs(title = "Distribution of Enrollment for Public and Private Universities",
  x = "Enrollment", y = "Density") + theme_minimal() + xlim(-5000,70000) + 
  scale_fill_manual(values = c("Public" = "blue", "Private" = "red")) +
  scale_color_manual(values = c("Public" = "blue", "Private" = "red"))
```

d. Any single variables of interest.
plotting the mean of in-state and out of state tuition for each university by university rank 
```{r}
college_rankings$Rank <- as.numeric(college_rankings$Rank)
ggplot(data = college_rankings, aes(x = Rank, y = Tuition.in.state)) + 
  geom_point() + geom_smooth(method = "loess") + scale_x_reverse() + labs(x = "Rank (Descending)",
  y = "In-State Tuition", title = "In-State Tuition as a Function of Rank")
```

e. Any other interesting plots you may find.
```{r}
college_rankings$Rank <- as.numeric(college_rankings$Rank)
ggplot(data = college_rankings, aes(x = Rank, y = Freshman.retention.rate)) + geom_point() + 
  geom_smooth(method = "loess") + scale_x_reverse() + labs(x = "Rank (Descending)", y = "In-State Tuition",
  title = "Freshman retention rate as a function of rank")
```























