---
title: "Untitled"
output: html_document
date: "2024-11-09"
---

1.1.1 Which dimension combination caused the issue? Explore the data and visualization to understand when the issue (possibly a significant number of failures in transactions ) happened and which combination of dimension (pmt, pg, bank and sub_type) has the impact. Tip: Identify the method to detect an anomaly in a metric across 4+ dimensions and apply that method to find the above.

```{r}
setwd("C:/Users/jake pc/Desktop/DS_Exam_2")
transactions <- read.csv(file="transactions(1).csv",header=TRUE)

transactions_original <- transactions
```

add column for failure % of each observation

```{r}
failure_percent <- numeric(nrow(transactions))
failure_percent <- (1 - transactions[,2]/transactions[,1])*100
transactions$failure_percent <- failure_percent

transactions <- transactions[order(as.POSIXct(transactions[,7], format = "%Y-%m-%d %H")),]
(unique_hours <- unique(transactions[,7]))

percentage_of_failed_transactions_per_hour <- numeric(length(unique_hours))
for (i in 1:72) {
  percentage_of_failed_transactions_per_hour[i] <- ( 1 - sum(transactions[transactions[,7] == unique_hours[i],2])/sum(transactions[transactions[,7] == unique_hours[i],1]))
}

failed_transactions_per_hour <- numeric(length(unique_hours))
for (i in 1:72) {
  percentage_of_failed_transactions_per_hour[i] <- ( 1 - sum(transactions[transactions[,7] == unique_hours[i],2]))
}

plot(x=seq(1,72,by=1),y=failed_transactions_per_hour,type="o")

plot(x=seq(1,72,by=1),y=percentage_of_failed_transactions_per_hour,type="o")
```



```{r}

transactions$weighted_failure_score <- transactions[,9] * transactions[,1] *100

test_pmt <- as.factor(transactions$pmt)
(levels(test_pmt))

test_pg <- as.factor(transactions$pg)
(levels(test_pg))

(transactions$bank[levels(as.factor(transactions$bank == 304))])

# Convert categorical features to numeric (using dummy encoding or integer encoding)
transactions$pmt <- as.numeric(as.factor(transactions$pmt))
transactions$pg <- as.numeric(as.factor(transactions$pg))
transactions$bank <- as.numeric(as.factor(transactions$bank))
transactions$sub_type <- as.numeric(as.factor(transactions$sub_type))

# Select relevant features for anomaly detection
features <- transactions[, c("weighted_failure_score", "pmt", "pg", "bank", "sub_type")]

# Calculate the Mahalanobis distance for each observation
center <- colMeans(features)
cov_matrix <- cov(features)
mahalanobis_distances <- mahalanobis(features, center, cov_matrix)

# Append Mahalanobis distances to the transactions dataframe
transactions$mahalanobis_score <- mahalanobis_distances

# Determine a threshold for high anomaly scores (e.g., 95th percentile)
threshold <- quantile(mahalanobis_distances, 0.999)
high_mahalanobis_transactions <- transactions[transactions$mahalanobis_score > threshold, ]

# Display the top anomalous transactions
head(high_mahalanobis_transactions, 100)



anamolous <- transactions_original[row.names(high_mahalanobis_transactions),]

anamolous
```

1.1.2 When did the issue happen? What is the starting hour and ending hour of the reported issue? Illustrate your answer with a plot

```{r}

```
