---
title: "Untitled"
output: html_document
date: "2024-11-19"
---

```{r}
setwd("/Users/jacobrichards/Desktop/DS_DA_Projects/Renewal_Response")
#setwd("C:/Users/jake pc/Desktop/DS_DA_Projects/Renewal_Response")
data <- read.csv(file="lease_renewal.csv",header=TRUE)
data
```

Important Note: Please do not build any kind of regression or classification model to address the questions below

What are the key drivers that explain whether a resident will renew the lease or not? In other words, what are the renewal promoters and detractors?

First lets divide the data set into tenants that renewed and those that did not.

```{r}
#rows that did renew 
RENEW <- data[data[,18] == 1, ]

#rows that did not renew 
NOT_RENEW <- data[data[,18] == 0, ]

#rent increase less than 10 
RENT_INCREASE_LT_TEN <- data[data[,2] == 0 & data[,3] == 0 & data[,4] == 0,]

#rent increased 10 percent 
RENT_INCREASE_10 <- data[data[,3] == 1,]

#rent increased 20 percent 
RENT_INCREASE_20 <- data[data[,4] == 1,]

#there was no rent change at all 
NO_RENT_INCREASE <- data[data[,2] == 1,]

RR <- matrix(nrow=2,ncol=2)

```

```{r}
library(fmsb)

RR <- matrix(nrow=2,ncol=2)

RR[1,1] <- length(intersect(row.names(RENT_INCREASE_LT_TEN),row.names(NOT_RENEW)))

RR[1,2] <- length(intersect(row.names(NO_RENT_INCREASE),row.names(NOT_RENEW)))

riskratio(RR[1,1], RR[1,2], nrow(RENT_INCREASE_LT_TEN), nrow(NO_RENT_INCREASE), conf.level=0.95, p.calc.by.independence=TRUE) 
```

```{r}
RR[1,1] <- length(intersect(row.names(RENT_INCREASE_10),row.names(NOT_RENEW)))

RR[1,2] <- length(intersect(row.names(NO_RENT_INCREASE),row.names(NOT_RENEW)))

riskratio(RR[1,1], RR[1,2], nrow(RENT_INCREASE_10), nrow(NO_RENT_INCREASE), conf.level=0.95, p.calc.by.independence=TRUE) 
```

```{r}
RR[1,1] <- length(intersect(row.names(RENT_INCREASE_20),row.names(NOT_RENEW)))

RR[1,2] <- length(intersect(row.names(NO_RENT_INCREASE),row.names(NOT_RENEW)))

riskratio(RR[1,1], RR[1,2], nrow(RENT_INCREASE_20), nrow(NO_RENT_INCREASE), conf.level=0.95, p.calc.by.independence=TRUE) 
```

```{r}
RR_lease_length <- matrix(nrow=2,ncol=2)

FIRST_LEASE <- data[data[,7]==1,]

#tenants that have renewed their lease 3 or more times. 
LEASE_GT_THIRD <- data[ data[,5]== 0 & data[,6]== 0 & data[,7]== 0, ]



RR[1,1] <- length(intersect(row.names(FIRST_LEASE),row.names(NOT_RENEW)))

RR[1,2] <- length(intersect(row.names(LEASE_GT_THIRD),row.names(NOT_RENEW)))


riskratio(RR[1,1], RR[1,2], nrow(FIRST_LEASE), nrow(LEASE_GT_THIRD), conf.level=0.95, p.calc.by.independence=TRUE) 
FIRST_LEASE

# 1.05 RR 

```

```{r}
SECOND_LEASE <- data[data[,5]==1,]

RR[1,1] <- length(intersect(row.names(SECOND_LEASE),row.names(NOT_RENEW)))

riskratio(RR[1,1], RR[1,2], nrow(SECOND_LEASE), nrow(LEASE_GT_THIRD), conf.level=0.95, p.calc.by.independence=TRUE) 
# 0.93
```

```{r}
THIRD_LEASE <- data[data[,6]==1,]

RR[1,1] <- length(intersect(row.names(THIRD_LEASE),row.names(NOT_RENEW)))

riskratio(RR[1,1], RR[1,2], nrow(THIRD_LEASE), nrow(LEASE_GT_THIRD), conf.level=0.95, p.calc.by.independence=TRUE) 

# 0.97 
```

append a column that indicated that the average age of the lease is over 60 if none of the other age categories are posative

```{r}


#of the age group under 24 only 9% renewed their lease. 
ratio_age_under_24_renewed <- sum(data[data[,8] == 1,18])/sum(data$age_range_under_24)
ratio_age_under_24_renewed


#24 to 29 22% renewed 
ratio_age_24_29_renewed <- sum(data[data[,9] == 1,18])/sum(data$age_range_24_29) 
ratio_age_24_29_renewed


# 30 to 39 66% renewed their lease 
ratio_age_30_39_renewed <- sum(data[data[,10] == 1,18])/sum(data$age_range_30_39)
ratio_age_30_39_renewed


# 40 to 49 
ratio_age_40_49_renewed <- sum(data[data[,11] == 1,18])/sum(data$age_range_40_49)
ratio_age_40_49_renewed

# 50 to 59 
ratio_age_50_59_renewed <- sum(data[data[,12] == 1,18])/sum(data$age_range_50_59)
ratio_age_50_59_renewed


ratio_age_50_59_renewed <- sum(data[data[,12] == 1,18])/sum(data$age_range_50_59)
ratio_age_50_59_renewed

ratio_age_60_renewed <- sum(data[data[,13] == 1,18])/sum(data$age_range_60)
ratio_age_60_renewed
```


24-49 had over 20% renewal rates 


